---
title: "BST 213 Homework 5"
author: "Matthew Moll"
output:
  html_document:
    toc: true
    toc_float: true
---

date: `r Sys.Date()`



## R setup
```{r setup}


knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 6)

library(tidyverse)
library(data.table)
library(knitr)
library(readxl)
library(moments)
library(epiR)
library(e1071)
library(lsmeans)
library(lmSupport)
library(ppcor)
library(tableone)
library(sjPlot)
library(stargazer)
library(car)
library(leaps)

```

## Load data
```{r load data}

rm(list = ls())

## define directory
analysisDir <- '/Users/matthewmoll/Documents/Fellowship/MPH/Fall2018/BST213_regression/homework/'
  
## Read in data
icu <- read_excel("/Users/matthewmoll/Documents/Fellowship/MPH/Fall2018/BST213_regression/homework/icu-2.xls")

```

## Data structure
```{r data structure}

head(icu)
str(icu)
sapply(icu, summary)


```

## Define variables

First, I will define these variables as continuous, categorical and binary, and then explore the simple relationships of each variable with the outcome (vital_status).

<BR>

```{r define variables}

removevars <- c('id')

## Define response variables
responseVars <- c('status')
responseVars

## Define explanatory variables
explanVars <- names(icu)[!names(icu) %in% c("status",removevars)]
explanVars

## List of quantitative variables
quantVars <- names(icu[!names(icu) %in% removevars &
                                  sapply(icu, function(x)
                                      length(levels(as.factor(x)))>7)])
quantVars


## List of binary variables
binVars <- names(icu[!names(icu) %in% removevars &
                                sapply(icu, function(x)
                                    length(levels(as.factor(x)))==2)])
binVars

## List of categorical variables

catVars <- names(icu[!names(icu) %in% c(removevars,binVars) &
                                sapply(icu, function(x)
                                    length(levels(as.factor(x)))<=7 & length(levels(as.factor(x)))>1)])
catVars


## Get Intersection between quantVars and explanVars and responseVars
explanVars.quant <- intersect(explanVars,quantVars)
responseVars.quant <- intersect(responseVars,quantVars)

## Get intersection between categorical and binary variables with explanatory and response vars
explanVars.cat <- intersect(explanVars,catVars)
explanVars.cat <- explanVars.cat[!explanVars.cat %in% c(binVars)]

responseVars.cat <- intersect(responseVars, catVars)
responseVars.cat <- responseVars.cat[!responseVars.cat %in% c(binVars)]


## Get binary explanatory and response variables
explanVars.bin <- intersect(explanVars, binVars)
responseVars.bin <- intersect(responseVars, binVars)

## View data ranges
str(icu)
sapply(icu, summary)

## turn binary and category variables into factors
varsToFactor <- c(catVars, binVars) 

icu[,varsToFactor] <- lapply(icu[,varsToFactor],function(x) as.factor(x))


```

<BR>

## Normality assessment
```{r normality assessment}

icu_quant <- as.data.frame(icu %>% dplyr::select(quantVars))


normalityfun <- function(dataset) {
  require(moments)
  require(e1071)
  
  print(paste0("Summary Statistics: ", names(dataset)[i]))
  print(paste0("Mean: ", mean(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Standard Deviation: ",sd(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Median: ", median(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Skewness: ", skewness(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Kurtosis: ", e1071::kurtosis(as.numeric(dataset[,i]), type = 2)))
  print(paste0("Shapiro-Wilk Test: ", shapiro.test(as.numeric(dataset[,i]))$p.value))
  print(" ")
  print(" ")
  
  qqnorm(as.numeric(dataset[,i]), main = paste0("Normal Q-Q plot: ",names(dataset)[i]))
  qqline(as.numeric(dataset[,i]), col = 2)
  
  
}


## Assess normality for continuous variables without stratifying for status

for(i in 1:length(quantVars)) {
  
  ## For alive subjects
    print(ggplot() + 
            geom_histogram(data = icu_quant, fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = paste0(quantVars[i]))
    
          )
  
    normalityfun(icu_quant)
  
}


## Now assess normality within groups defined by vital status

icu_quant.alive <- icu %>% filter(status == 0) %>% dplyr::select(quantVars) %>% as.data.frame()
icu_quant.dead <- icu %>% filter(status == 1) %>% dplyr::select(quantVars) %>% as.data.frame()

for(i in 1:length(quantVars)) {
  
  ## For alive subjects
    print(ggplot() + 
            geom_histogram(data = icu_quant.alive, 
                           fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = paste0("Alive:  ",quantVars[i]))
    
          )
  
    normalityfun(icu_quant.alive)


    ## For dead subjects
    
     print(ggplot() + 
            geom_histogram(data = icu_quant.dead, 
                           fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = paste0("Dead:  ",quantVars[i]))
    
          )
  
    normalityfun(icu_quant.dead)
  
}



```

<BR>

None of these are normally distributed, but I will invoke central limit theorem for sys and heart. Age is not normally distributed, so I will use nonparametric testing for age.

```{r}

nonParamVars <- c('age')

```

<BR>

## Binary vs continuous

In this section, I will use boxplots and t-tests to compare the continuous explanatory variables stratified by the binary outcome, vital_status.

```{r visualize boxplots, eval = T}

## Before using t-test, need to figure out whether to use equal or unqual variance.

for(d in 1:length(explanVars.quant[!explanVars.quant %in% c(nonParamVars)])) {
  
  print(" ")
  print(paste0("Variable:  ",explanVars.quant[!explanVars.quant %in% c(nonParamVars)][d]))
  print(" ")
  
  f.test <- 
  var.test(icu_quant.alive[,explanVars.quant[!explanVars.quant %in% c(nonParamVars)][d]], 
                  
              icu_quant.dead[,explanVars.quant[!explanVars.quant %in% c(nonParamVars)][d]], 
              alternative = "two.sided")
  
  print(f.test)

}


for(i in 1:length(responseVars)) {
  
  for(j in 1:length(explanVars.quant[!explanVars.quant %in% c(nonParamVars)])) {
    
    ## First do parametric testing for normally distributed variables
    
     print(paste0("Response var: ", responseVars[i]))
     print(paste0("Explan var: ", explanVars.quant[!explanVars.quant %in% c(nonParamVars)][j]))
     
     diff <- 
       t.test(icu_quant.alive[,explanVars.quant[!explanVars.quant %in% c(nonParamVars)][j]], 
                  
              icu_quant.dead[,explanVars.quant[!explanVars.quant %in% c(nonParamVars)][j]], 
              alternative = "two.sided", var.equal = 
                ifelse(explanVars.quant[!explanVars.quant %in% c(nonParamVars)][j] == "sys",FALSE,TRUE))
     
     print(paste0("p-value: ", signif(diff$p.value,5)))
     
     print(" ")
     print(" ")
     print(" ")
     print(" ")
     print(" ")
     
       
        print(ggplot(data = icu, mapping = aes(
          x = get(responseVars[i]),
          y = get(explanVars.quant[!explanVars.quant %in% c(nonParamVars)][j]), 
          group = get(responseVars[i]))) +
          geom_boxplot() +
          xlab(responseVars[i]) +
          ylab(explanVars.quant[!explanVars.quant %in% c(nonParamVars)][j]) +
          labs(subtitle = paste0("p-value: ", signif(diff$p.value,5),", ", paste0("mean x: ",
               signif(diff$estimate[1], 5)),", ", paste0("mean y: ", signif(diff$estimate[2], 5)))))
    
       
  
        }
  
   ## Do non-parametric testing for non-normal variables
        for(k in 1:length(nonParamVars)) {
 
          print(paste0("Response var: ", responseVars[i]))
          print(paste0("Explan var: ", nonParamVars[k]))
     
          diff <- wilcox.test(icu_quant.alive[,nonParamVars[k]], 
                              icu_quant.dead[,nonParamVars[k]])
     
     
          print(paste0("p-value: ", signif(diff$p.value,5)))
     
     
          print(" ")
          print(" ")
          print(" ")
          print(" ")
          print(" ")
     
       
          print(ggplot(data = icu, mapping = aes(
          x = get(responseVars[i]),
          y = get(nonParamVars[k]), 
          group = get(responseVars[i]))) +
          geom_boxplot() +
          xlab(responseVars[i]) +
          ylab(nonParamVars[k]) +
          labs(subtitle = paste0("p-value: ", signif(diff$p.value,5))))

        }
  }



```

## Binary X Binary
```{r binary X binary}

## What are possible binary exposures?
explanVars.bin

explanVars.bin2 <- explanVars.bin 

## What are binary outcomes of interest?
responseVars.bin

for (n in 1:length(responseVars.bin)) {
  
  for (o in 1:length(explanVars.bin2)) {
    
    exposureframe <- na.omit(as.data.frame(icu[,c(explanVars.bin2[o],responseVars.bin[n])]))
    
    exposuretable <- table(exposureframe[,explanVars.bin2[o]], exposureframe[,responseVars.bin[n]])
    print(" ")
    print(" ")
    print(paste0("Exposure: ", explanVars.bin2[o]))
    print(paste0("Outcome: ", responseVars.bin[n]))
    print(" ")
    print(" ")
 
      
    epitable <-  epi.2by2(dat = exposuretable, method = "cross.sectional",
          conf.level = 0.95, units = 100, homogeneity = "breslow.day",
         outcome = "as.columns")
    
    print(epitable)
    
    print(" ")

    ## Fisher test
    
    print(
    fisher.test(x = as.data.frame(icu)[,explanVars.bin2[o]], 
            y = as.data.frame(icu)[,responseVars.bin[n]], 
            alternative = "two.sided", conf.int = TRUE)
    )
    
    }
  
}

```

## Binary X Categorical
```{r binary X categorical}

for(i in 1:length(responseVars.bin)) {
  
  for (j in 1:length(explanVars.cat)) {
    
    print(" ")
    print(" ")
    
    print(paste0("Explan Var:  ",explanVars.cat[j]))
    print(paste0("Response Var:  ",responseVars.bin[i]))
    
    print(" ")
    print(" ")
    
    print(
    fisher.test(x = as.data.frame(icu)[,explanVars.cat[j]], 
            y = as.data.frame(icu)[,responseVars.bin[i]], 
            alternative = "two.sided", conf.int = TRUE)
    )

  }
}




```



## Session Info
```{r session info}

sessionInfo()

```