---
title: "BST 213 Homework 6"
author: "Matthew Moll"
output:
  html_document:
    toc: true
    toc_float: true
---

date: `r Sys.Date()`



## R setup
```{r setup}


knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 6)

library(tidyverse)
library(data.table)
library(knitr)
library(readxl)
library(moments)
library(epiR)
library(e1071)
library(lsmeans)
library(lmSupport)
library(ppcor)
library(tableone)
library(sjPlot)
library(stargazer)
library(car)
library(leaps)

```

## Load data
```{r load data}

rm(list = ls())

## define directory
analysisDir <- '/Users/matthewmoll/Documents/Fellowship/MPH/Fall2018/BST213_regression/homework/'
  
## Read in data
icu <- read_excel("/Users/matthewmoll/Documents/Fellowship/MPH/Fall2018/BST213_regression/homework/icu-2.xls")

```

## Data structure
```{r data structure}

head(icu)
str(icu)
sapply(icu, summary)


```

## Define variables

For this homework assignment, we are assigned the task of logistic regression for age (continuous), renal failure (binary), and race (categorical) for the outcome status. 

<BR>

```{r define variables}


## Define response variables
responseVars <- c('status')
responseVars

## Define explanatory variables
explanVars <- names(icu)[grep(paste0(c("age","renal","race"), collapse = "|"), names(icu), ignore.case = T)]
explanVars


## Variables to remove
removevars <- names(icu)[!names(icu) %in% c(explanVars,'status')]

## List of quantitative variables
quantVars <- names(icu[!names(icu) %in% removevars &
                                  sapply(icu, function(x)
                                      length(levels(as.factor(x)))>7)])
quantVars


## List of binary variables
binVars <- names(icu[!names(icu) %in% removevars &
                                sapply(icu, function(x)
                                    length(levels(as.factor(x)))==2)])
binVars

## List of categorical variables

catVars <- names(icu[!names(icu) %in% c(removevars,binVars) &
                                sapply(icu, function(x)
                                    length(levels(as.factor(x)))<=7 & length(levels(as.factor(x)))>1)])
catVars


## Get Intersection between quantVars and explanVars and responseVars
explanVars.quant <- intersect(explanVars,quantVars)
responseVars.quant <- intersect(responseVars,quantVars)

## Get intersection between categorical and binary variables with explanatory and response vars
explanVars.cat <- intersect(explanVars,catVars)
explanVars.cat <- explanVars.cat[!explanVars.cat %in% c(binVars)]

responseVars.cat <- intersect(responseVars, catVars)
responseVars.cat <- responseVars.cat[!responseVars.cat %in% c(binVars)]


## Get binary explanatory and response variables
explanVars.bin <- intersect(explanVars, binVars)
responseVars.bin <- intersect(responseVars, binVars)

## View data ranges
str(icu)
sapply(icu, summary)

## turn binary and category variables into factors
varsToFactor <- c(catVars, binVars) 

icu[,varsToFactor] <- lapply(icu[,varsToFactor],function(x) as.factor(x))

## define race category 1 (white) as the reference group
icu$race <- relevel(icu$race, ref = "1")

## For the sake of simplicity, create subset of icu for this analysis
icu <- icu %>% dplyr::select(responseVars,explanVars)


```

<BR>

## Normality assessment
```{r normality assessment}

icu_quant <- as.data.frame(icu %>% dplyr::select(quantVars))


normalityfun <- function(dataset) {
  require(moments)
  require(e1071)
  
  print(paste0("Summary Statistics: ", names(dataset)[i]))
  print(paste0("Mean: ", mean(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Standard Deviation: ",sd(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Median: ", median(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Skewness: ", skewness(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Kurtosis: ", e1071::kurtosis(as.numeric(dataset[,i]), type = 2)))
  print(paste0("Shapiro-Wilk Test: ", shapiro.test(as.numeric(dataset[,i]))$p.value))
  print(" ")
  print(" ")
  
  qqnorm(as.numeric(dataset[,i]), main = paste0("Normal Q-Q plot: ",names(dataset)[i]))
  qqline(as.numeric(dataset[,i]), col = 2)
  
  
}


## Assess normality for continuous variables without stratifying for status

for(i in 1:length(quantVars)) {
  
  ## For alive subjects
    print(ggplot() + 
            geom_histogram(data = icu_quant, fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = paste0(quantVars[i]))
    
          )
  
    normalityfun(icu_quant)
  
}


## Now assess normality within groups defined by vital status

icu_quant.alive <- icu %>% filter(status == 0) %>% dplyr::select(quantVars) %>% as.data.frame()
icu_quant.dead <- icu %>% filter(status == 1) %>% dplyr::select(quantVars) %>% as.data.frame()

for(i in 1:length(quantVars)) {
  
  ## For alive subjects
    print(ggplot() + 
            geom_histogram(data = icu_quant.alive, 
                           fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = paste0("Alive:  ",quantVars[i]))
    
          )
  
    normalityfun(icu_quant.alive)


    ## For dead subjects
    
     print(ggplot() + 
            geom_histogram(data = icu_quant.dead, 
                           fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = paste0("Dead:  ",quantVars[i]))
    
          )
  
    normalityfun(icu_quant.dead)
  
}



```

<BR>

None of these are normally distributed, but I will invoke central limit theorem for sys and heart. Age is not normally distributed, so I will use nonparametric testing for age.

```{r}

nonParamVars <- c('age')

```

<BR>

## Convert continuous variable to categorical
<BR>
Age can be converted into a categorical variable, or we can add a squared term. I think for a reader, it is easier to conceptualize categories of age and their relationship to the outcome of death.
<BR>

```{r}

## Split age into categories - will use 10 year increments
agegroups <- seq(10,100,10)

icu <- icu %>% mutate(agecat = ifelse(age >= agegroups[1] & age < agegroups[2], 1, 
                               ifelse(age >= agegroups[2] & age < agegroups[3], 2, 
                                      ifelse(age >= agegroups[3] & age < agegroups[4], 3, 
                                             ifelse(age >= agegroups[4] & age < agegroups[5], 4, 
                                                    ifelse(age >= agegroups[5] & age < agegroups[6], 5,
                                                           ifelse(age >= agegroups[6] & age < agegroups[7], 6,
                                                                  ifelse(age >= agegroups[7] & age < agegroups[8], 7,
                                                                      ifelse(age >= agegroups[8] & age < agegroups[9], 8, ifelse(age >= agegroups[9] & age < agegroups[10], 9, NA))))))))))

## Make a factor
icu$agecat <- as.factor(icu$agecat)

## add to explanVars

explanVars <- c(explanVars,'agecat')

```


## Univariate logistic regression
```{r}

for(i in 1:length(explanVars[!explanVars %in% c('age')])) {
  
  print(paste0("Variable:  ",explanVars[!explanVars %in% c('age')][i]))
  print(" ")
  
  logMod <- paste0("status~",explanVars[!explanVars %in% c('age')][i])

  logReg <- glm(as.formula(logMod), data = icu, family = "binomial")

  print(summary(logReg))
  
  ## Calculate odds ratios and 95% CI

  for(j in 2:length(summary(logReg)$coefficients[,1])) {
    
    print(" ")
    print(paste0("Odds ratio for ", 
                 names(summary(logReg)$coefficients[,1])[j],":  ", 
                 signif(exp(summary(logReg)$coefficients[j,1]),5), " (", 
                 signif((exp(summary(logReg)$coefficients[j,1]) - 1.96*summary(logReg)$coefficients[j,2]),5), 
                 ", ", 
                 signif((exp(summary(logReg)$coefficients[j,1]) + 1.96*summary(logReg)$coefficients[j,2]),5), 
                 ")"))
    # print(" ")
    # print(paste0("LCI for ", 
    #              names(summary(logReg)$coefficients[,1])[j],":  ", 
    #              signif((exp(summary(logReg)$coefficients[j,1]) - 1.96*summary(logReg)$coefficients[j,2]),5)))
    # print(" ")
    # print(paste0("UCI for ", 
    #              names(summary(logReg)$coefficients[,1])[j],":  ", 
    #              signif((exp(summary(logReg)$coefficients[j,1]) + 1.96*summary(logReg)$coefficients[j,2]),5)))
    
  }
  print(" ")
}


```

## Multivariable logistic regression
```{r}


logMod <- paste0("status~",paste0(explanVars[!explanVars %in% c('age')], collapse = "+"))

logReg <- glm(as.formula(logMod), data = icu, family = "binomial")

print(summary(logReg))

for(j in 2:length(summary(logReg)$coefficients[,1])) {
    
    print(" ")
    print(paste0("Odds ratio for ", 
                 names(summary(logReg)$coefficients[,1])[j],":  ", 
                 signif(exp(summary(logReg)$coefficients[j,1]),5), " (", 
                 signif((exp(summary(logReg)$coefficients[j,1]) - 1.96*summary(logReg)$coefficients[j,2]),5), 
                 ", ", 
                 signif((exp(summary(logReg)$coefficients[j,1]) + 1.96*summary(logReg)$coefficients[j,2]),5), 
                 ")"))
    # print(" ")
    # print(paste0("LCI for ", 
    #              names(summary(logReg)$coefficients[,1])[j],":  ", 
    #              signif((exp(summary(logReg)$coefficients[j,1]) - 1.96*summary(logReg)$coefficients[j,2]),5)))
    # print(" ")
    # print(paste0("UCI for ", 
    #              names(summary(logReg)$coefficients[,1])[j],":  ", 
    #              signif((exp(summary(logReg)$coefficients[j,1]) + 1.96*summary(logReg)$coefficients[j,2]),5)))
    
  }
  
print(" ")

```


## Session Info
```{r session info}

sessionInfo()

```