---
title: "BST 213 Midterm"
author: "Matthew Moll"
output:
  html_document:
    toc: true
    toc_float: true
---

date: `r Sys.Date()`



## R setup
```{r setup}


knitr::opts_chunk$set(echo = TRUE, fig.width = 9, fig.height = 9)

library(tidyverse)
library(data.table)
library(knitr)
library(readxl)
library(moments)
library(epiR)
library(e1071)
library(lsmeans)
library(lmSupport)
library(ppcor)
library(tableone)
library(sjPlot)
library(stargazer)
library(car)
library(leaps)

```

## Load data
```{r load data}

rm(list = ls())

## define directory
analysisDir <- '/Users/matthewmoll/Documents/Fellowship/MPH/Fall2018/BST213_regression/homework/'
  
## Read in data
cost <- read_excel("/Users/matthewmoll/Documents/Fellowship/MPH/Fall2018/BST213_regression/homework/cost.xls")

```

## Data structure
```{r data structure}

head(cost)
str(cost)
sapply(cost, summary)


```

## Define variables

The goal for this assignment is to build a linear regression model using the priniciples discussed in class, and then write it into a methods and results section of a manuscript.

<BR>

```{r define variables}

removevars <- c()

## Define response variables
responseVars <- c('outcosts2')
responseVars

## Define explanatory variables
explanVars <- names(cost)[!names(cost) %in% c("outcosts2")]
explanVars

## List of quantitative variables
quantVars <- names(cost[!names(cost) %in% removevars &
                                  sapply(cost, function(x)
                                      length(levels(as.factor(x)))>7)])
quantVars


## List of binary variables
binVars <- names(cost[!names(cost) %in% removevars &
                                sapply(cost, function(x)
                                    length(levels(as.factor(x)))==2)])
binVars

## List of categorical variables

catVars <- names(cost[!names(cost) %in% c(removevars,binVars) &
                                sapply(cost, function(x)
                                    length(levels(as.factor(x)))<=7 & length(levels(as.factor(x)))>1)])
catVars


## Get Intersection between quantVars and explanVars and responseVars
explanVars.quant <- intersect(explanVars,quantVars)
responseVars.quant <- intersect(responseVars,quantVars)

## Get intersection between categorical and binary variables with explanatory and response vars
explanVars.cat <- intersect(explanVars,catVars)
explanVars.cat <- explanVars.cat[!explanVars.cat %in% c(binVars)]

responseVars.cat <- intersect(responseVars, catVars)
responseVars.cat <- responseVars.cat[!responseVars.cat %in% c(binVars)]


## Get binary explanatory and response variables
explanVars.bin <- intersect(explanVars, binVars)
responseVars.bin <- intersect(responseVars, binVars)

## View data ranges
str(cost)
sapply(cost, summary)

```

<BR>

## Normality assessment
```{r normality assessment}

cost_quant <- as.data.frame(cost %>% dplyr::select(quantVars))


normalityfun <- function(dataset) {
  require(moments)
  require(e1071)
  
  print(paste0("Summary Statistics: ", names(dataset)[i]))
  print(paste0("Mean: ", mean(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Standard Deviation: ",sd(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Median: ", median(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Skewness: ", skewness(as.numeric(dataset[,i]), na.rm = T)))
  print(paste0("Kurtosis: ", e1071::kurtosis(as.numeric(dataset[,i]), type = 2)))
  print(paste0("Shapiro-Wilk Test: ", shapiro.test(as.numeric(dataset[,i]))$p.value))
  print(" ")
  print(" ")
  
  qqnorm(as.numeric(dataset[,i]), main = paste0("Normal Q-Q plot: ",names(dataset)[i]))
  qqline(as.numeric(dataset[,i]), col = 2)
  
  
}



for(i in 1:length(quantVars)) {
  
    print(ggplot() + 
            geom_histogram(data = cost_quant, fill = "blue", alpha = 0.5, mapping = 
                             aes(x = get(quantVars[i]))) +
                xlab(quantVars[i]) + labs(title = "Histogram")
    
  )
  
    normalityfun(cost_quant)
  
}



```

<BR>

The only variable that looks close to normal is patient age, which we may be able to use parametric testing by invoking the central limit theorem. Otherwise, we will report median and interquartile ranges.

<BR>

## Missingness
```{r missingness}

## assess for and remove > 10% missingness
## Set threshold
thr <- 0.1*length(cost$outcosts2)

sapply(cost, function(x) ifelse(sum(is.na(x)) > thr, 1, 0))

sapply(cost,function(x) sum(is.na(x)))

## Create complete dataset of phenotypic variables
cost <- na.omit(cost)

```

No variables were removed for missingness. However, ssi has the greatest missingness.

## Table 1: Demographics
```{r table 1}

## Add labels to table

labelled::var_label(cost$outcosts2) <- "Outpatient costs ($)"
labelled::var_label(cost$ptage) <- "Age in years"
labelled::var_label(cost$ssi) <- "Somatic Symptom Index (No. (%))"
labelled::var_label(cost$panic) <- "Panic (No. (%))"
labelled::var_label(cost$anxiety) <- "Anxiety (No. (%))"
labelled::var_label(cost$married) <- "Married (No. (%))"
labelled::var_label(cost$female) <- "Female (No. (%))"
labelled::var_label(cost$educat) <- "Education Level (No. (%))"
labelled::var_label(cost$iadl) <- "IADL score"
labelled::var_label(cost$social) <- "Social Activities Score"
labelled::var_label(cost$racecat) <- "Race (No. (%))"
labelled::var_label(cost$whiteleycat) <- "Whiteley Category"
labelled::var_label(cost$charlson) <- "Charlson index"
labelled::var_label(cost$depcat) <- "Depression Category (No. (%))"




## Group variables

demovars <- names(cost)
demovars

varsToFactor <- c(catVars, binVars) 

nonNormalVars <- names(cost_quant)[!names(cost_quant) %in% c("ptage")]

## Table one

tableOne <- CreateTableOne(vars = demovars, data = cost, factorVars = varsToFactor, testNonNormal = nonNormalVars)
tableOne

## Save as csv and word document
tableOne_out <- print(tableOne,quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabel = TRUE, nonnormal = nonNormalVars)


write.csv(tableOne_out, file = paste0(analysisDir,"TableOne.csv"))
tableOne_out <- fread(paste0(analysisDir, "TableOne.csv"))

colnames(tableOne_out) <- c("","")

tab_df(tableOne_out, file = paste0(analysisDir, "TableOne.doc"))


```


## Scatterplots and correlation coefficients
```{r scatterpolots and correlation coefficients}

## Scatterplots and correlation coefficients

for (i in 1:length(explanVars.quant)) {
  
  for (j in 1:length(responseVars.quant)) {
 
     print(paste0("Explanatory var: ", explanVars.quant[i]))
     print(paste0("Response var: ", responseVars.quant[j]))
     
     cormod <- cor.test(as.data.frame(cost)[,explanVars.quant[i]],
                                    as.data.frame(cost)[,responseVars.quant[j]], 
                        method = "spearman")
     
     print(paste0("Spearman Correlation Coefficient: ", signif(cormod$estimate,5)))
     print(paste0("p-value: ", signif(cormod$p.value,5)))
     
     print(" ")
     print(" ")
  
     print(ggplot(data = cost) + 
             geom_point(mapping = aes(x = get(explanVars.quant[i]), 
                                      y = get(responseVars.quant[j]))) +
             geom_smooth(mapping = aes(x = get(explanVars.quant[i]), 
                                       y = get(responseVars.quant[j])), method = 'lm') +
             xlab(explanVars.quant[i]) + ylab(responseVars.quant[j]) +
               labs(subtitle =
                      paste0(paste0("r-value: ", signif(cormod$estimate,5)),", ",
                             paste0("p-value: ", signif(cormod$p.value,5))))) 
     
     ## plot residuals
     print(ggplot(data = lm(get(responseVars.quant[j])~get(explanVars.quant[i]), data = cost)) + 
             geom_point(aes(x = .fitted, y = .resid)) + 
             labs(title = paste0("Residuals vs fitted: ",explanVars.quant[i]))) # fitted vs residuals
     
    print(ggplot(data = lm(get(responseVars.quant[j])~get(explanVars.quant[i]), data = cost)) + 
            geom_histogram(aes(x = .resid)) + 
            labs(title = paste0("Residuals histogram: ",explanVars.quant[i])))       
     

    }
}



```

<BR>


## All vars: Scatterplots and correlation coefficients
```{r scatterpolots and correlation coefficients for all vars}

## Scatterplots and correlation coefficients

for (i in 1:length(explanVars)) {
  
  for (j in 1:length(responseVars.quant)) {
 
     print(paste0("Explanatory var: ", explanVars[i]))
     print(paste0("Response var: ", responseVars.quant[j]))
     
     cormod <- cor.test(as.data.frame(cost)[,explanVars[i]],
                                    as.data.frame(cost)[,responseVars.quant[j]], 
                        method = "spearman")
     
     print(paste0("Spearman Correlation Coefficient: ", signif(cormod$estimate,5)))
     print(paste0("p-value: ", signif(cormod$p.value,5)))
     
     print(" ")
     print(" ")
  
     print(ggplot(data = cost) + 
             geom_point(mapping = aes(x = get(explanVars[i]), 
                                      y = get(responseVars.quant[j]))) +
             geom_smooth(mapping = aes(x = get(explanVars[i]), 
                                       y = get(responseVars.quant[j])), method = 'lm') +
             xlab(explanVars[i]) + ylab(responseVars.quant[j]) +
               labs(subtitle =
                      paste0(paste0("r-value: ", signif(cormod$estimate,5)),", ",
                             paste0("p-value: ", signif(cormod$p.value,5))))) 
     
     ## plot residuals
     print(ggplot(data = lm(get(responseVars.quant[j])~get(explanVars[i]), data = cost)) + 
             geom_point(aes(x = .fitted, y = .resid)) + 
             labs(title = paste0("Residuals vs fitted: ",explanVars[i]))) # fitted vs residuals
     
    print(ggplot(data = lm(get(responseVars.quant[j])~get(explanVars[i]), data = cost)) + 
            geom_histogram(aes(x = .resid)) + 
            labs(title = paste0("Residuals histogram: ",explanVars[i])))       
     

    }
}



```


## Variable selection
```{r variable selection}


# 
# FS <- regsubsets(as.formula(LinEq), data = cost, method = "forward", force.in = c("depcat","charlson"))
# 
# str(summary(FS))
# str(FS)
# 
# summary(FS)$rss

set.seed(123)

null <- lm(outcosts2~1, data = cost)
full <- lm(outcosts2~., data = cost)

FS <- step(null, scope=list(lower=null, upper=full), direction="forward")

## View best model
summary(FS)

## Create vector of selected variables
selectedVars <- rownames(summary(FS)$coefficients)
selectedVars <- selectedVars[!selectedVars %in% c("(Intercept)")]
selectedVars

summary(FS)$coefficients


## do by p-value
# OLS.Mod <- paste0("outcosts2~depcat+charlson+",paste0(names(cost)[!names(cost) %in% c("charlson","depcat","outcosts2")], collapse = "+"))
# OLS <- ols(as.formula(OLS.Mod), data = cost)
# 
# FS.p <- fastbw(OLS, rule = "p", sls = 0.05, force = 4)

```

## Multicollinearity: variance inflation factors
```{r VIF}

LinEq <- paste0("outcosts2~",paste0(selectedVars, collapse = "+"))
LinEq

LinMod <- lm(as.formula(LinEq), data = cost)
summary(LinMod)

vif(LinMod)

## Remove if VIF > 10 - none in this case

## Now create table for adjusted coefficients
coefTable <- 
  data.frame(
    variable = 
      rownames(summary(LinMod)$coefficients)[2:length(summary(LinMod)$coefficients[,1])], 
    coefficient = 
      as.numeric(summary(LinMod)$coefficients[2:length(summary(LinMod)$coefficients[,1]),1]), 
    LCI = 
      as.numeric(confint(LinMod)[2:length(summary(LinMod)$coefficients[,1]),1]), 
    UCI = as.numeric(confint(LinMod)[2:length(summary(LinMod)$coefficients[,1]),2]), 
    pvalue = 
               signif(as.numeric(summary(LinMod)$coefficients[2:length(summary(LinMod)$coefficients[,1]),4]),9))

kable(coefTable)


## Now remove SSI which is likely an intermediary variable

LinEq.int <- paste0("outcosts2~",paste0(selectedVars[!selectedVars %in% c("ssi")], collapse = "+"))
LinEq.int

LinMod.int <- lm(as.formula(LinEq), data = cost)
summary(LinMod)

vif(LinMod.int)


```

## Univariate OLS
```{r univariate OLS}


coefTable.uni <- data.frame(variable = character(),coefficient = numeric(), LCI = numeric(), UCI = numeric(), pvalue = numeric())

for(i in 1:length(selectedVars)) {
  
  
  LinEq.uni <- paste0("outcosts2~",selectedVars[i])
  LinEq.uni

  LinMod.uni <- lm(as.formula(LinEq.uni), data = cost)

  coefTable.uni <- rbind(coefTable.uni, data.frame(variable = selectedVars[i], 
                            coefficient = as.numeric(summary(LinMod.uni)$coefficients[2,1]), 
                            LCI = confint(LinMod.uni)[2,1], 
                            UCI = confint(LinMod.uni)[2,2], 
                            pvalue = signif(summary(LinMod.uni)$coefficients[2,4], 9)
                       ))

}


kable(coefTable.uni)

```

## Table 2: coefficients
```{r table 2 coefficients}

table2 <- bind_cols(coefTable.uni,coefTable)
table2 <- as.data.frame(sapply(table2 %>% dplyr::select(-variable,-variable1), function(x) signif(x, 5)))
kable(table2)


## if p < 0.0001
table2[,c("pvalue","pvalue1")] <- sapply(table2[,c("pvalue","pvalue1")], function(x) ifelse(x < 0.0001, "< 0.0001",x))


## Create vector rownames and extract labels
variableNames <- selectedVars

variableNameLabels <- vector(length = length(variableNames))

for(y in 1:length(variableNameLabels)) {
  
  variableNameLabels[y] <- labelled::var_label(cost[,variableNames[y]])
}


## Collapse 95% CIs into cell with hazard ratios

table2 <- table2 %>% mutate(variable = variableNameLabels, 
                       Coef1 = paste0(coefficient," (",LCI,",",UCI,")"), 
                       Coef2 = paste0(coefficient1," (",LCI1,",",UCI1,")")) %>% dplyr::select(variable,Coef1,pvalue,Coef2,pvalue1)

colnames(table2) <- c("Variable", "Unadjusted Coefficient (95% CI)","p-value","Adjusted Coefficient (95% CI)","p-value")

kable(table2)

tab_df(table2, file = paste0(analysisDir,"Table2.doc"))


```


## Effect modification
```{r effect modification}

## Evaluate for effect modification by adding an interaction term
cost <- cost %>% mutate(chdep = charlson*depcat)
labelled::var_label(cost$chdep) <- 'Charlson*Depression Category Interaction Term' # label this new term

## Retrain the model with the new interaction term
LinEq <- paste0("outcosts2~",paste0(c(selectedVars[!selectedVars %in% c("ssi")],"chdep"), collapse = "+"))
LinEq

LinMod <- lm(as.formula(LinEq), data = cost)
summary(LinMod)

vif(LinMod)

```


## Figure 1: plot of outpatient costs vs. depression category grouped by depression category
```{r figure 1}

## plot line of best fit for outpatient costs vs charlson grouped by depression category

ggplot(data = cost[,c("outcosts2",selectedVars)], 
       aes(x = charlson, y = outcosts2,group = as.factor(depcat), color = as.factor(depcat))) + 
  geom_smooth(method = 'lm') + 
  ggtitle(" ") + xlab("Charlson Index") + ylab("Outpatient Costs") + 
  labs(color = "Depression \nCategory\n") + 
  scale_color_manual(labels = c("Major","Minor","None"), values = c("red","blue","dark green"))
  
## Save file
pdf(paste0(analysisDir,"Figure1.pdf"))

ggplot(data = cost[,c("outcosts2",selectedVars)], 
       aes(x = charlson, y = outcosts2,group = as.factor(depcat), color = as.factor(depcat))) + 
  geom_smooth(method = 'lm') + 
  ggtitle(" ") + xlab("Charlson Index") + ylab("Outpatient Costs") + 
  labs(color = "Depression \nCategory\n") + 
  scale_color_manual(labels = c("Major","Minor","None"), values = c("red","blue","dark green"))
 

dev.off()

```




## Session Info
```{r session info}

sessionInfo()

```